Validation of Visual Statistical Inference, Applied to Linear Models

### 2012-06-03 paper_jasa.tex
generation of lineup

\begin{figure}[htp]
%\begin{figurehere}
   \centering
       \includegraphics[width=0.95\textwidth]{test_category.pdf}
       \caption{Lineup plot ($m=20$) using side-by-side boxplots for testing $H_0: \beta_k=0$. One of these plots is the plot of the actual data, and the remaining are null plots, produced by simulating data from a null model that assumes $H_0$ is true. Which plot is the most different from the others, in the sense that there is the the largest shift or location difference between the boxplots? (The position of the actual data plot is provided in Section \ref{sec:category}.)}
       \label{fig:test_category}
\end{figure}
%\end{figurehere}

@@

 Sampling Distribution & $f_{T(y)}(t); $\begin{minipage}[h]{2.5cm} \begin{center} \scalebox{0.55}{\includegraphics{stat_mathematical_test.pdf}} \end{center} \end{minipage} & $f_{T(y)}(t); $ \begin{minipage}[h]{2.5cm} \begin{center} \scalebox{0.32}{\includegraphics{test_category_small.pdf}} \end{center} \end{minipage} \\
 & \begin{minipage}[h]{2.5cm} \begin{center} \scalebox{0.2}{\includegraphics{down_arrow.pdf}} \end{center} \end{minipage} & \begin{minipage}[h]{2.5cm} \begin{center} \scalebox{0.2}{\includegraphics{down_arrow.pdf}} \end{center} \end{minipage} \\
 Reject $H_0$ if & actual $T$ is extreme & actual plot is identifiable \\
\hline 
\end{tabular}
\label{tbl:compare}
\end{table*}

@@

\begin{table*}[hbtp] 
\caption{Test Statistics for Testing Hypothesis Related to Model $Y_i = \beta_0 + \beta_1 X_{i1} + \beta_1 X_{i2} + \beta_3 X_{i1}X_{i2} + ... + \epsilon_i  $ } 
\centering 
\begin{tabular}{m{0.5cm}m{3cm}m{2.5cm}m{2.6cm}m{5.5cm}} 
\hline\hline 
Case & Null Hypothesis & Statistic & Test Statistic & Description \\ [0.5ex] % inserts table %heading 
\hline 
1 & $H_0: \beta_0=0$ & Scatter plot & \begin{minipage}[t]{2cm}\begin{center}  	\scalebox{0.1}{\includegraphics{stat_intercept.pdf}} \end{center} \end{minipage} & Scatter plot with least square line overlaid. For lineup plot, we simulate data from fitted null model. \\ 
2 & $H_0: \beta_k=0$ & Residual plot & \begin{minipage}[t]{2cm}\begin{center}  	\scalebox{0.1}{\includegraphics{stat_bet_p.pdf}} \end{center} \end{minipage} & Residual vs $X_k$ plots. For lineup plot, we simulate data from normal with mean 0 variance $\hat{\sigma}^2$. \\ 
3 & $H_0: \beta_k=0$ (for categorical $X_k$) & Box plots of residuals & \begin{minipage}[t]{2cm}\begin{center}  	\scalebox{0.45}{\includegraphics{stat_category.pdf}} \end{center} \end{minipage} & Box plot of residuals grouped by category of $X_k$. For lineup plot, we simulate data from normal with mean 0 variance $\hat{\sigma}^2$. \\
4 & $H_0: \beta_k=0$ (interaction with categorical $X_k$) & Scatter plot & \begin{minipage}[t]{2cm}\begin{center}  \scalebox{0.11}{\includegraphics{stat_interection.pdf}} \end{center} \end{minipage} & Scatter plot with least square lines of each category overlaid. For lineup plot, we simulate data from fitted null model.\\[1ex] % [1ex] adds vertical space 
5 & $H_0: X$ Linear & Residual Plot & \begin{minipage}[t]{2cm}\begin{center}  	\scalebox{0.1}{\includegraphics{stat_nonlinear.pdf}} \end{center} \end{minipage} & Residual vs predictor plots with loess smoother overlaid. For lineup plot, we simulate residual data from normal with mean 0 variance $\hat{\sigma}^2$. \\ 
6 & $H_0: \sigma^2=\sigma^2_0$ & Box plot & \begin{minipage}[t]{2cm}\begin{center}  \scalebox{0.15}{\includegraphics{stat_sigma_box.pdf}} \end{center} \end{minipage} & Box plot of standardized residual divided by  $\sigma^2_0$. For lineup plot, we simulate data from standard normal. \\ 				
7 & $H_0: \rho_{X,Y|Z}=\rho$ & Scatter Plot & \begin{minipage}[t]{2cm}\begin{center}  	\scalebox{0.1}{\includegraphics{stat_intercept.pdf}} \end{center} \end{minipage} & Scatter plot of Residuals obtained by fitting partial regression. For lineup plot, we simulate data (mean 0 and variance 1) with specific correlation $\rho$. \\ 
8 & $H_0:$ Model Fits & Histogram & \begin{minipage}[t]{2cm}\begin{center}  \scalebox{0.1}{\includegraphics{stat_goodness_simple.pdf}} \end{center} \end{minipage} & Histogram of the response data. For lineup plot, we simulate data from fitted model. \\[1ex] % [1ex] adds vertical space 
9 & Special case $p=1$ $H_0: \rho_{X,Y} =\rho$ & Scatter plot & \begin{minipage}[t]{2cm}\begin{center}  \scalebox{0.1}{\includegraphics{stat_intercept.pdf}} \end{center} \end{minipage} & Scatter plot with least square line overlaid For lineup plot, we simulate data with correlation $\rho$.\\			
\hline 
\end{tabular} 
\label{tbl:stat_multiple} 
\end{table*}

@@

@@

@@

%For a given proportion $p$ we want to have margin of error (ME) to be 0.1. Thus we have $ME =1.96 \sqrt{ \frac 1 n p(1-p)} \le 0.05$ which gives us the estimation of minimum number of evaluations $$n_p \geq \frac{p(1-p)}{(0.05/1.96)^2}.$$ Now from the theoretical (UMP) power curve for each parameter combinations shown in table \ref{tbl:experiment_params} we know the power(say $p$) and thus we can estimate the number of evaluations required for that specific combinations of parameter. We collect data such that each of the parameter combinations in the table \ref{tbl:experiment_params} has approximately the estimated sample size that we calculated here.
  Figure \ref{fig:test_category} is an example of one of these lineups. The visual test statistic is a side-by-side boxplot.

@@

\green{Did you in fact bin like this???} For added control, to ensure a signal in the simulated actual data a blocking structure was used to filter data sets. A 1000 sets were generated for each parameter combination and the traditional $t$-statistic and $p$-value associated with $H_0: \beta_2=0$ were calculated. The 3 replicates were drawn from each of three blocks of $p$-values: (0.0-$q_{33}$), ($q_{33}$-$q_{66}$), ($q_{66}$-1) where $q_i$ is the $i$th percentile in the distribution of the $p$-values. Additional control was applied to the 19 null plots. Because the distribution of these $p$-values should follow a Uniform(0,1) distribution, data sets were binned on this range by $p$-value, and a data set was randomly selected from each bin.

@@

This experiment is designed for the same regression model (Equation \ref{multi}) with $p=1$ and covariate $X_1$ being continuous. Data is simulated with two sample sizes ($n=100, 300$) and two standard deviations of the error ($\sigma=5, 12$). The different values for slope parameter that are used to generate data are shown in Table \ref{tbl:experiment_params}. We arbitrarily fix $\beta_0 = 6$ while generating actual data. For each combination of parameter, we generate at least three different actual data sets constituting a total of 65 lineups.

\green{Fill in details of lineups, probably need a figure with an example lineup, and how lineups were assigned to subjects.}

>>>>

\begin{figure}[htp]
%\begin{figurehere}
   \centering
       \includegraphics[width=0.95\textwidth]{lineup_category.pdf}
       \caption{Lineup plot ($m=20$) using side-by-side boxplots for testing $H_0: \beta_k=0$. One of these plots is the plot of the actual data, and the remaining are null plots, produced by simulating data from a null model that assumes $H_0$ is true. Which plot is the most different from the others, in the sense that there is the the largest shift or location difference between the boxplots? (The position of the actual data plot is provided in Section \ref{sec:category}.)}
       \label{fig:test_category}
\end{figure}
%\end{figurehere}

@@

 Sampling Distribution & $f_{T(y)}(t); $\begin{minipage}[h]{2.5cm} \begin{center} \scalebox{0.55}{\includegraphics{stat_mathematical_test.pdf}} \end{center} \end{minipage} & $f_{T(y)}(t); $ \begin{minipage}[h]{2.5cm} \begin{center} \scalebox{0.32}{\includegraphics{lineup_category_small.pdf}} \end{center} \end{minipage} \\
 & \begin{minipage}[h]{2.5cm} \begin{center} \scalebox{0.2}{\includegraphics{down_arrow.pdf}} \end{center} \end{minipage} & \begin{minipage}[h]{2.5cm} \begin{center} \scalebox{0.2}{\includegraphics{down_arrow.pdf}} \end{center} \end{minipage} \\
 Reject $H_0$ if & actual $T$ is extreme & actual plot is identifiable \\
\hline 
\end{tabular}
\label{tbl:compare}
\end{table*}

@@

\begin{table*}[hbtp] 
\caption{Test Statistics for Testing Hypothesis Related to Model $Y_i = \beta_0 + \beta_1 X_{i1} + \beta_1 X_{i2} + \beta_3 X_{i1}X_{i2} + ... + \epsilon_i  $ } 
\centering 
\begin{tabular}{m{0.5cm}m{3cm}m{2.5cm}m{2.6cm}m{5.5cm}} 
\hline\hline 
Case & Null Hypothesis & Statistic & Test Statistic & Description \\ [0.5ex] % inserts table %heading 
\hline 
1 & $H_0: \beta_0=0$ & Scatter plot & \begin{minipage}[t]{2cm}\begin{center}  	\scalebox{0.1}{\includegraphics{stat_intercept.pdf}} \end{center} \end{minipage} & Scatter plot with least square line overlaid. For lineup plot, we simulate data from fitted null model. \\ 
2 & $H_0: \beta_k=0$ & Scatter plot & \begin{minipage}[t]{2cm}\begin{center}  	\scalebox{0.45}{\includegraphics{stat_beta_k.pdf}} \end{center} \end{minipage} & Scatter plot with least square line overlaid. For lineup plot, we simulate data from fitted null model. \\ 
3 & $H_0: \beta_k=0$ (for categorical $X_k$) & Box plot & \begin{minipage}[t]{2cm}\begin{center}  	\scalebox{0.45}{\includegraphics{stat_category.pdf}} \end{center} \end{minipage} & Box plot of residuals grouped by category of $X_k$. For lineup plot, we simulate data from normal with mean 0 variance $\hat{\sigma}^2$. \\
4 & $H_0: \beta_k=0$ (interaction with categorical $X_k$) & Scatter plot & \begin{minipage}[t]{2cm}\begin{center}  \scalebox{0.11}{\includegraphics{stat_interection.pdf}} \end{center} \end{minipage} & Scatter plot with least square lines of each category overlaid. For lineup plot, we simulate data from fitted null model.\\[1ex] % [1ex] adds vertical space 
5 & $H_0: X$ Linear & Residual Plot & \begin{minipage}[t]{2cm}\begin{center}  	\scalebox{0.1}{\includegraphics{stat_nonlinear.pdf}} \end{center} \end{minipage} & Residual vs predictor plots with loess smoother overlaid. For lineup plot, we simulate residual data from normal with mean 0 variance $\hat{\sigma}^2$. \\ 
6 & $H_0: \sigma^2=\sigma^2_0$ & Box plot & \begin{minipage}[t]{2cm}\begin{center}  \scalebox{0.15}{\includegraphics{stat_sigma_box.pdf}} \end{center} \end{minipage} & Box plot of standardized residual divided by  $\sigma^2_0$. For lineup plot, we simulate data from standard normal. \\ 				
7 & $H_0: \rho_{X,Y|Z}=\rho$ & Scatter Plot & \begin{minipage}[t]{2cm}\begin{center}  	\scalebox{0.1}{\includegraphics{stat_intercept.pdf}} \end{center} \end{minipage} & Scatter plot of Residuals obtained by fitting partial regression. For lineup plot, we simulate data (mean 0 and variance 1) with specific correlation $\rho$. \\ 
8 & $H_0:$ Model Fits & Histogram & \begin{minipage}[t]{2cm}\begin{center}  \scalebox{0.1}{\includegraphics{stat_goodness_simple.pdf}} \end{center} \end{minipage} & Histogram of the response data. For lineup plot, we simulate data from fitted model. \\[1ex] % [1ex] adds vertical space 
9 & Special case $p=1$ $H_0: \rho_{X,Y} =\rho$ & Scatter plot & \begin{minipage}[t]{2cm}\begin{center}  \scalebox{0.1}{\includegraphics{stat_intercept.pdf}} \end{center} \end{minipage} & Scatter plot with least square line overlaid For lineup plot, we simulate data with correlation $\rho$.\\			
\hline 
\end{tabular} 
\label{tbl:stat_multiple} 
\end{table*}

@@

{\bf Plotting visual statistic:} For a given data set we fit the null model and obtain the residuals.  Using these residuals the visual statistics for each of the experiments were generated as described in table \ref{tbl:stat_multiple}. Actual data plot and null plots in a lineup are obtained from actual data set and null data set respectively. The procedure for obtaining actual data sets are described in the following sections.

@@

{\bf Steps to obtain lineup}
\begin{enumerate}
	\item For a given parameter setting obtain actual data set
	\item Fit the null model to the actual data and obtain residuals and other parameter estimates ($\hat{\beta}$ and $\hat{\sigma}$)
	\item Generate actual data plot using the procedure as described in table \ref{tbl:stat_multiple}
	\item Generate 19 null (residual) data sets from $N(0, {\hat{\sigma}}^2$) and obtain 19 null plots
	\item Finally randomly place 19 null plots and one actual plot in a layout (4X5) to get a lineup 
\end{enumerate}

 Figure \ref{fig:test_category} is an example of one of these lineups. The visual test statistic is a side-by-side boxplot. The actual data plot is number 13 which was generated for $n$=300, $\beta_2$=15 and $\sigma$=12.

@@

%For a given proportion $p$ we want to have margin of error (ME) to be 0.1. Thus we have $ME =1.96 \sqrt{ \frac 1 n p(1-p)} \le 0.05$ which gives us the estimation of minimum number of evaluations $$n_p \geq \frac{p(1-p)}{(0.05/1.96)^2}.$$ Now from the theoretical (UMP) power curve for each parameter combinations shown in table \ref{tbl:experiment_params} we know the power(say $p$) and thus we can estimate the number of evaluations required for that specific combinations of parameter. We collect data such that each of the parameter combinations in the table \ref{tbl:experiment_params} has approximately the estimated sample size that we calculated here.

@@

\green{Did you in fact bin like this???}\red{For each parameter combination a set of 1000 actual data are generated and their associated p-values were computed. This produces an empirical distribution of p-value under alternative hypothesis. Finally, three data sets that have 16th, 50th and 84th quantile of the p-values were taken. Similar procedure was used to generate 19 null plots based on the empirical distribution of p-value under null hypothesis}. For added control, to ensure a signal in the simulated actual data a blocking structure was used to filter data sets. A 1000 sets were generated for each parameter combination and the traditional $t$-statistic and $p$-value associated with $H_0: \beta_2=0$ were calculated. The 3 replicates were drawn from each of three blocks of $p$-values: (0.0-$q_{33}$), ($q_{33}$-$q_{66}$), ($q_{66}$-1) where $q_i$ is the $i$th percentile in the distribution of the $p$-values. Additional control was applied to the 19 null plots. Because the distribution of these $p$-values should follow a Uniform(0,1) distribution, data sets were binned on this range by $p$-value, and a data set was randomly selected from each bin.

@@

This experiment is designed for the same regression model (Equation \ref{multi}) with $p=1$ and covariate $X_1$ being continuous. $X_1$ is simulated from standard normal(0,1). Data is simulated with two sample sizes ($n=100, 300$) and two standard deviations of the error ($\sigma=5, 12$). The different values for slope parameter that are used to generate data are shown in Table \ref{tbl:experiment_params}. We arbitrarily fix $\beta_0 = 6$ while generating actual data. For each combination of parameter, we generate at least three different actual data sets constituting a total of 65 lineups.

{\bf Steps to obtain lineup}
\begin{enumerate}
	\item For a given parameter setting obtain actual data set
	\item Fit the null model to the actual data and obtain residuals and other parameter estimates ($\hat{\beta}$ and $\hat{\sigma}$)
	\item Generate actual data plot (a scatter plot of $Y$ vs $X_1$ a least square line overlaid)
	\item Generate 19 null data sets from $N(X \hat{\beta}, {\hat{\sigma}}^2$) and obtain 19 null plots
	\item Finally randomly place 19 null plots and one actual plot in a layout (4X5) to get a lineup 
\end{enumerate}

%\green{Fill in details of lineups, probably need a figure with an example lineup, and how lineups were assigned to subjects.}

\begin{figure}[htp]
%\begin{figurehere}
   \centering
       \includegraphics[width=0.95\textwidth]{lineup_continuous.pdf}
       \caption{Lineup plot ($m=20$) using scatter plots for testing $H_0: \beta_k=0$ where covariate $X_k$ is continuous. One of these plots is the plot of the actual data, and the remaining are null plots, produced by simulating data from a null model that assumes $H_0$ is true. Which plot is the most different from the others, in the sense that there is the steepest slope?}
       \label{fig:test_continous}
\end{figure}
